Early Warning Signals and the Prosecutor's Fallacy

### 2012-09-03 prosecutor/fallacy.tex
updated discussion

\usepackage{endfloat} % Move floats to the end of the document
\renewcommand{\listoffigures}{} % but supress these lists
\renewcommand{\listoftables}{} % supress these lists
% compile without figures?
\newif\ifhavefigures
%\havefigurestrue
\newif\ifnohavefigures % allows alternate condition
\nohavefigurestrue     % turn on the alternate condition uncommenting this

@@

\begin{figure}
  \begin{center}
    \subfigure[\textbf{True criminal} A bifurcation
      drives a transition into the undesirable state]{
      \ifhavefigures
%        \includegraphics[width=.4\textwidth]{figure1a}
      \fi
      \ifnohavefigures
        \hspace{.4\textwidth}
      \fi
      \label{fig:bifurcation}
    } 
    \subfigure[\textbf{Innocent suspect} A
    transition to the alternative stable state has occurred purely by
    chance.]{
    \ifhavefigures
%      \includegraphics[width=.4\textwidth]{figure1b}
    \fi
    \ifnohavefigures
      \hspace{.4\textwidth}
    \fi
    \label{fig:fallacy} }
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
  state versus potential in two cases.  Potential is the negative integral
  of the mean of dynamical equation.  The potential function gives an
  intuitive picture of the stability of a system by imagining the curve as
  a surface on which a ball is free to bounce across -- wells correspond
  to stable points and peaks to unstable points.  Early warning signals
  are aimed at detecting systems which are slowly moving towards a tipping
  point or bifurcation, illustrated in the successively darker potential
  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
  a simulation under this process shows the state of the system as the
  potential moves towards the bifurcation point. The state of the system
  in the simulation is given by the horizontal axis, while the vertical
  dimension is used to show time on the right-hand axis.  Some stable
  systems, panel~\subref{fig:fallacy}, may experience chance transitions
  to an alternate stable state without the presence of this slow change
  in the shape of the potential well which tips the system off to the left.
  An example of the trajectory of such a chance transition is
  plotted over the potential.  While such events may be unlikely over
  short time-scales, selecting historical examples of systems that
  have experienced sudden transitions increases the probability of
  accidentally using chance events depicted in panel~\subref{fig:fallacy}
  to test methods designed to detect the kind of process depicted in
  panel~\subref{fig:bifurcation}.} \label{fig:1}
\end{figure}

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulate a system that is not driven towards a bifurcation such as
in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
examining historical events is a valid way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The model can be represented as a continuous time birth-death process where births and deaths are Poisson events which depend on the current density with rates given by

\begin{align}
    b(n) &= \frac{e K n^2}{n^2 + h^2}, \label{master} \\
    d(n) &= e n + a, 
\end{align}

@@

%\begin{align}
%  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
%    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
%    d_n &= e n + a,
%\end{align}

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ is a non-parametric measure of rank correlation frequently used to identify an increasing trend ($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011}, defined as $\tau \def \tfrac{(\text{number of concordant pairs}) - (\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$ observations.  \footnote{A pair of observations $(x_i,y_i)$ and $(x_j, y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and $y_i < y_j$ and discordant otherwise; equalties excepted.}  $\tau$ takes values in $(-1, 1)$.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.   
%The common statistical significance test used for Kendall's $\tau$
%assumes independence of the variables to achieve a normal distribution,
%this isn't satisfied for indicators coming from a time series and
%computed with a sliding window.  For a more robust comparison, 
We compare the distribution of $\tau$ from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state.  To avoid an effect of sample size the time series are all chosen
to be the same length.  
%As these examples have been selected from a system
%that is not approaching a bifurcation, an indicator that does not suffer
%from the bias of Prosecutor's fallacy should have the same distribution
%in both cases.   
Annotated, parallelised code to replicate the analysis
and produce the figures is provided in the appendix.

\begin{figure}
  \begin{center}
    \ifhavefigures
%    \includegraphics[width=6in]{figure2}
    \fi
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

It seems tempting to argue that this bias towards positive detection in
historical examples is not problematic -- each of these systems did
indeed collapse, so the increased probability of exhibiting warning
signals could be taken as a successful detection.  Unfortunately this
is not the case. At the moment the forecast is made, these systems are
no more likely to crash than their counterparts from the sample that did
not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  A closer look at the patterns involved shows why
common indicators such as autocorrelation and variance can be misleading.

@@

@@

\subsection{Importance of experimental approaches}
The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem embodied is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases from some larger historical sample.  
%We note as well that the Prosecutors Fallacy 
%may arise in other contexts where we search historical data for
%examples containing some feature of interest.

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
manuscript.

>>>>

\usepackage{endfloat} % Move floats to the end of the document
\renewcommand{\listoffigures}{} % but supress these lists
\renewcommand{\listoftables}{} % supress these lists
% compile without figures?
\newif\ifhavefigures
\havefigurestrue
\newif\ifnohavefigures % allows alternate condition
%\nohavefigurestrue     % turn on the alternate condition uncommenting this

@@

\begin{figure}
  \begin{center}
      \ifhavefigures
        \includegraphics[width=\textwidth]{Figure1}
      \fi
      \ifnohavefigures
        \hspace{\textwidth}
      \fi
      \label{fig:bifurcation}
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. 
\textbf{(a)} Plot of the model functions shown in Eq~\eqref{master}
with parameters $a=180$, $K=500$, $e=.5$, and $h=200$.  When the death
rate is higher than the birth rate, the system dynamics drive the state
(population size) to smaller values.  When birth rate is higher, the
system moves right, as indicated by the arrows.\textbf{(b)} The potential
energy is given by the negative integral of $b(n)-d(n)$, shown in the
lower plot.   The potential function gives an intuitive picture of
the stability of a system by imagining the curve as a surface on which
a ball is free to bounce across -- wells correspond to stable points
and peaks to unstable points. While most trajectories remain near the
stable well, some transition out merely by chance.  An example of such
a trajectory is shown in the top panel, in which time increases along
the vertical axis. Though initially oscillating around the stable state,
a chance excursion carries it beyond the threshold (vertical dotted line).
Such chance trajectories can produce the statistical patterns as observed
in true critical transitions seen in panel \textbf{(c)}: Early warning
signals  are aimed at detecting systems which are slowly moving towards
a tipping point or bifurcation, illustrated in the successive curves
(detiorating and critical). Top panel: An example trajectory from a
simulation under this process shows the state of the system as the
potential moves towards the bifurcation point.
} \label{fig:1}
\end{figure}

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulate a system that is not driven towards a bifurcation such as
in Fig~ref{fig:1}(b).  This simulation approach allows us to determine whether
examining historical events is a valid way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The model can be represented as a continuous time birth-death process where births and deaths are Poisson events which depend on the current density with rates given by

\begin{align}
    b(n) &= \frac{ K n^2}{n^2 + h^2}, \label{master} \\
    d(n) &= e n + a, 
\end{align}

@@

%\begin{align}
%  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
%    b_n &= \frac{K n^2}{n^2 + h^2}, \\
%    d_n &= e n + a,
%\end{align}

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ is a non-parametric measure
of rank correlation frequently used to identify an increasing trend
($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011},
defined as $\tau \def \tfrac{(\text{number of concordant pairs}) -
(\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$
observations.  \footnote{A pair of observations $(x_i,y_i)$ and $(x_j,
y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and
$y_i < y_j$ and discordant otherwise; equalities excepted.}  $\tau$ takes
values in $(-1, 1)$.  The distribution of $\tau$ values observed across
these replicates is shown in Figure~\ref{fig:indicator}.  We compare
the distribution of $\tau$ from all the simulations to the distribution
conditioned on experiencing a chance transition to the alternative
stable state.  To avoid an effect of sample size the time series are
all chosen to be the same length.

To demonstrate the effect we observe is not unique to models with Allee
effects, we provide an example of the effect arising in a discrete-time
model with two non-zero stable states adapted from~\citep{May1977},
\begin{equation}
X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{s ^ Q + H ^ Q} \right). \label{May1977}
\end{equation}
Annotated, parallel code to replicate the analysis presented here is
also provided in the appendix.

\begin{figure}
  \begin{center}
    \ifhavefigures
    \includegraphics[width=6in]{figure2}
    \fi
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

It seems tempting to argue that this bias towards positive detection
in historical examples is not problematic -- each of these systems did
indeed collapse, so the increased probability of exhibiting warning
signals could be taken as a successful detection.  Unfortunately this is
not the case. At the moment the forecast is made, these systems are not
likely to transition, since they experience a strong pull towards the
original stable state.  A closer look at the patterns involved shows why
common indicators such as autocorrelation and variance can be misleading.

@@

\subsection{ The truncation of observations }

If we had a complete knowledge of the system dynamics, then we could
eliminate the bias we observe here since the bias arises from the
transient branch of the trajectory that crosses the threshold, and
if the system were truncated at the minimum of the potential then the
effect we emphasize here would not appear.  But, it is not possible to
truncate the system in any practical application. The precise location of
the minimum of the potential which is the location of the deterministic
equilibrium is unknown. Moreover, under the hypothesis that the system is
approaching a critical transition, the location of the minimum potential
moves so it cannot easily be estimated by previous observations, (see
Figure 1c where the equilibrium point moves in the direction of the
transition). Thus it is neither practical nor desirable to suggest that
historical time series can be used by following a simple truncation rule
that avoids the branch of a trajectory crossing the threshold to another
basin of attraction.  Exactly where a particular study will choose
to truncate such a trajectory will necessarily be arbitrary without an
underlying model of the process.  Frequently this is done by removing
the very steep, monotonic branch of the trajectory expected once the
system crosses the unstable threshold.  Such an approach corresponds
with our choice of termination and produces the bias we discuss here.

The examples of Figure 1, though only single replicates, may be useful
in illustrating these issues.  Figure 1c, top panel shows a sample
trajectory of a system with a parameter shift, while 1b shows a trajectory
without a shift.  Both trajectories become more highly autocorrelated
and higher variance near the end of the time series (time increases
on the y axis in Figure 1).  The part of the time series following
the critical transition shows a fast and monotonic trajectory to the
unstable trajectory, and would usually be excluded by an analysis for
warning signals in advance of the transition.  No such clear pattern
exists prior to the transition in Figure 1b.   An alternative proposal
to terminate the trajectory in panel B earlier would also risk decreasing
the signal seen in panel c, and would be inconsistent with the application
of warning signals in the forecasting context, where there would be no
such truncation.

@@

\subsection{Importance of experimental approaches} The problem we
highlight ultimately stems from the difficulty of having only a single
realization with which to examine a complex problem.  The only way
to deal with this problem embodied is through replication, as can
be done in an experimental system in laboratory manipulations such
as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole
lake ecosystems in~\citet{Carpenter2011}.  Experimental procedures avoid
the hazard of the Prosecutor's fallacy by generating a complete sample
of replicates, rather than selecting a subset of cases from some larger
historical sample.

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of
Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
manuscript, and also P. Ditlevsen and an anonymous reviewer for their 
comments.

\section{References}%bibliography
\bibliography{fallacy}
